\documentclass[12pt]{article}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{amsbsy}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{array}
\usepackage{booktabs}
\usepackage{epsfig}
\usepackage{fancyhdr}
\usepackage{float}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{ifthen}
\usepackage{times}
\usepackage{xspace}
\usepackage[colorlinks]{hyperref}
\usepackage[numbers,sort&compress]{natbib}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

\usepackage{boldtensors}
\usepackage{microtype}

% Allow line breaks inside \texttt{...} at underscores and after "::"
% so long code identifiers (e.g. CapModel_Def.hpp, PHAL::AlbanyTraits::Jacobian)
% do not overrun the right margin.
\setlength{\emergencystretch}{3em}
\makeatletter
\def\BreakableUnderscore{\leavevmode\nobreak\hskip\z@{\char`\_}\nobreak\hskip\z@\allowbreak}
\makeatother
\AtBeginDocument{\def\_{\BreakableUnderscore}}
\newcommand{\origcolcol}{::}
\newcommand{\bcc}{::\allowbreak}

\DeclareMathOperator{\trace}{tr}
\DeclareMathOperator{\dev}{dev}
\DeclareMathOperator{\Grad}{Grad}
\DeclareMathOperator{\sym}{sym}
\newcommand{\transpose}[1]{{#1}^{\mathrm{T}}}
\newcommand{\Reals}{\mathbb{R}}

% Model-specific symbols
\newcommand{\stress}{~\sigma}
\newcommand{\kirch}{~\tau}
\newcommand{\backstress}{~\alpha}
\newcommand{\relstress}{~\xi}
\newcommand{\estrain}{{~\varepsilon^e}}
\newcommand{\pstrain}{{~\varepsilon^p}}
\newcommand{\strainincr}{\Delta~\varepsilon}
\newcommand{\Fdef}{~F}
\newcommand{\Fplas}{{~F^p}}
\newcommand{\bee}{{~b^e}}
\newcommand{\Celast}{{~{\mathsf C}^e}}
\newcommand{\ident}{~I}
\newcommand{\sdev}{~s}
\newcommand{\halpha}{~h^{\alpha}}

\begin{document}

\setlength{\headheight}{15pt}
\headsep = 4pt
\pagestyle{fancyplain}

\lhead
[\fancyplain{}{\emph{LCM Documentation}}]
{\fancyplain{}{\emph{LCM Documentation}}}

\rhead
[\fancyplain{}{\emph{Cap Plasticity}}]
{\fancyplain{}{\emph{Cap Plasticity}}}

\title{
  Cap Plasticity in LCM:\\
  Reference Formulation, Integration, and Verification}

\author{
  Alejandro Mota$^1$\thanks{Email: amota@sandia.gov}\\
  \\
  $^1$Mechanics of Materials Department\\
  Sandia National Laboratories\\
  Livermore CA 94550, USA\\
}

\date{\today}

\maketitle

\begin{abstract}

  LCM implements a single cap-plasticity constitutive model,
  \texttt{CapModel} (material-database name \texttt{Cap}): the
  three-invariant isotropic/kinematic hardening cap model of the
  Sandia GeoModel lineage
  \citep{Fossum.Fredrich:2000,Fossum.Brannon:2004,Foster.etal:2005},
  in the form presented in Section~3 of \citet{Sun.Chen.Ostien:2014}.
  This document states the formulation as implemented, including the
  conventions that the published papers leave ambiguous and that were
  settled against the original sources and the LAME GeoModel reference
  implementation; documents the adaptive substepped explicit
  integration; describes the logarithmic-strain / Kirchhoff-stress
  kinematics used for finite deformation; summarizes why no
  algorithmic tangent is ever hand-coded (Sacado automatic
  differentiation produces all of them); and describes the
  verification of the implementation against an independent reference
  implementation, which reproduces the kernel to machine precision
  over full load histories.
  \\
  \\
  Keywords: Cap plasticity, three-invariant yield surface, kinematic
  hardening, pore collapse, adaptive substepping, logarithmic strain,
  Kirchhoff stress, automatic differentiation, verification, LCM.

\end{abstract}

\section{Scope, Sources, and Conventions}
\label{sec:scope}

\texttt{CapModel} lives in
\texttt{src/LCM/models/CapModel.\{hpp,cpp\}} and
\texttt{CapModel\_Def.hpp} and is selected in the material database
with \texttt{Model Name: Cap}. It supports small-strain and
finite-deformation kinematics through the boolean material parameter
\texttt{Finite Deformation} (Section~\ref{sec:finite}); the
constitutive integrator is identical in both cases.

A fully implicit variant (\texttt{CapImplicit}, a 13-unknown local
backward-Euler solve following \citealt{Foster.etal:2005}) existed in
LCM until 2026 and was removed: it was unmaintained, had no test
coverage, and its principal classical advantage --- the consistent
algorithmic tangent --- carries little weight in LCM, where automatic
differentiation produces the exact tangent of whatever integrator
actually runs (Section~\ref{sec:ad}). The accuracy concern that
remains for an explicit scheme, large strain increments, is addressed
by adaptive substepping (Section~\ref{sec:integration}). The model's
own authors made the same choice, moving from the implicit
formulation of \citet{Foster.etal:2005} to the refined explicit
scheme of \citet{Sun.Chen.Ostien:2014} for production work.

\paragraph{Sources and their precedence.}
The implementation follows
\citet[Section~3]{Sun.Chen.Ostien:2014}. Where that paper is
ambiguous, in error, or silent, the conventions are settled by, in
order: the original model papers
\citep{Foster.etal:2005,Regueiro.Foster:2011}, and the LAME GeoModel
reference implementation (\texttt{iso\_geomodel\_model.F}, the
production Sandia code documented in \citealt{Fossum.Brannon:2004}).
Each such arbitration is called out where it appears below. Two
errata in \citet{Sun.Chen.Ostien:2014} as printed are relevant:
\begin{enumerate}
\item Algorithm~1, Step~2 prints the stress update as
      $\stress_{n+1} = \stress^{\mathrm{tr}}
       - \Delta\gamma\,\partial g/\partial\stress$, omitting the
      elastic tangent. The $\Delta\gamma$ derivation in the same
      paper, and Algorithm~2, both require
      $\stress_{n+1} = \stress^{\mathrm{tr}}
       - \Delta\gamma\,\Celast\!:\!\partial g/\partial\stress$,
      which is what LCM implements.
\item Algorithm~2's normal-correction fallback mixes $\Celast$ into
      the stress update while omitting it from the denominator
      $\tilde\chi = \partial_\sigma f : \partial_\sigma f$, which is
      dimensionally inconsistent. LCM implements the consistent
      \citet{Sloan.Abbo.Sheng:2001} form: a pure geometric projection
      with no elastic tangent in either place
      (Section~\ref{sec:integration}).
\end{enumerate}

\paragraph{Sign convention.}
Tension positive, compression negative, throughout. Under
confinement $I_1 = \trace\stress < 0$; the cap sits on the
compressive (negative-$I_1$) side, and ``the cap expands'' means its
hydrostat intercept moves toward $-\infty$.

\section{Physical Overview}
\label{sec:physics}

Geomaterials exhibit two inelastic mechanisms that the yield surface
must represent simultaneously. At low to moderate pressure the
dominant mechanism is frictional sliding and growth of microcracks:
shear resistance grows with confinement, producing a pressure-dependent
\emph{shear failure envelope} that, unlike a Drucker--Prager cone,
saturates once the cracks are squeezed shut. At high pressure a
porous material yields under hydrostatic load alone by irreversible
\emph{pore collapse}; representing this requires closing the yield
surface on the compressive end with a \emph{cap}, which moves outward
as porosity is crushed away --- the isotropic hardening of the model.
Shear stress assists pore collapse, so the cap is an ellipse rather
than a flat cutoff. Directionality of frictional sliding produces a
strong Bauschinger effect, represented by a deviatoric back stress
that translates the yield surface up to, but never beyond, the fixed
failure envelope. Finally, geomaterials are weaker in triaxial
extension than in triaxial compression, which brings the third stress
invariant into the yield function. The classical two-surface cap
models \citep[e.g.][]{Fossum.Fredrich:2000} join envelope and cap at
a corner; the present formulation is a single $C^1$ surface, which is
the property that makes the explicit integration well behaved.

\section{Kinematics in the Small-Strain Path}
\label{sec:smallstrain}

The small-strain path additively decomposes the strain rate
\citep[Eq.~47]{Sun.Chen.Ostien:2014},
\begin{equation}
  \dot{~\varepsilon} \;=\; \dot{\estrain} + \dot{\pstrain},
\end{equation}
and uses isotropic linear elasticity
\citep[Eqs.~49--50]{Sun.Chen.Ostien:2014},
\begin{equation}
  \dot{\stress} \;=\; \Celast : \dot{\estrain}
    \;=\; \Celast : (\dot{~\varepsilon} - \dot{\pstrain}),
  \qquad
  \Celast \;=\; \lambda~I \otimes ~I + 2\mu ~{\mathbb I},
\end{equation}
with $\lambda,\mu$ the Lam\'e constants and $~{\mathbb I}$ the
symmetric fourth-order identity. Given state
$\{\stress_n, \backstress_n, \kappa_n\}$ and strain increment
$\strainincr$ from the surrounding global Newton iteration, the trial
elastic stress is
\begin{equation}
  \stress^{\mathrm{tr}} \;=\; \stress_n + \Celast : \strainincr .
  \label{eq:trial_small}
\end{equation}

\section{Invariants, Lode Coefficient, and Yield Surface}
\label{sec:yield}

Kinematic hardening is introduced through the deviatoric back stress
$\backstress$ via the relative stress
\begin{equation}
  \relstress \;:=\; \stress - \backstress,
\end{equation}
with invariants
\begin{align}
  I_1        &= \trace \relstress, \\
  J_2^{\xi}  &= \tfrac{1}{2}\, \sdev : \sdev,
                \qquad \sdev \;=\; \relstress - \tfrac{1}{3} I_1 ~I, \\
  J_3^{\xi}  &= \det \sdev .
\end{align}
Because $\backstress$ is deviatoric,
$\trace\relstress = \trace\stress$, so the $\xi$ superscript is used
on $J_2, J_3$ only.

\paragraph{Lode coefficient.}
The difference in strength between triaxial compression and triaxial
extension enters through
\citep[Eqs.~34--36]{Sun.Chen.Ostien:2014}
\begin{equation}
  \Gamma(\beta) \;=\;
    \tfrac{1}{2}\left[
      1 - \sin 3\beta
    + \frac{1}{\psi}\left(1 + \sin 3\beta\right)
    \right],
  \qquad
  \sin 3\beta \;=\; -\frac{3\sqrt{3}\,J_3^\xi}{2\,(J_2^\xi)^{3/2}},
  \label{eq:Gamma}
\end{equation}
where $\psi \le 1$ is the ratio of triaxial-extension to
triaxial-compression strength. In triaxial compression $\Gamma = 1$;
in triaxial extension $\Gamma = 1/\psi$, and since $\Gamma^2$
multiplies $J_2^\xi$ in the yield function, yield in extension occurs
at $\psi$ times the compressive strength. The code sets
$\Gamma \equiv 1$ when $\psi = 0$ or $J_2^\xi = 0$ to avoid the
singularity on the hydrostat. With $\psi = 1$ the Lode dependence is
inactive and all $J_3$ terms drop out of the stress gradients ---
a fact relevant to the implementation history
(Appendix~\ref{sec:history}).

\paragraph{Shear failure surface, cap, and cap position.}
The pressure-dependent shear failure surface and the smooth
elliptical hardening cap
\citep[Eqs.~32, 37--38]{Sun.Chen.Ostien:2014} are
\begin{align}
  F_f(I_1) &= A - C \exp(D\,I_1) - \theta\,I_1,
             \label{eq:Ff}\\
  F_c(I_1,\kappa) &=
    1 - H(\kappa - I_1)\,\frac{(I_1 - \kappa)^2}{[X(\kappa) - \kappa]^2},
             \label{eq:Fc}\\
  X(\kappa) &= \kappa - R\,F_f(\kappa),
             \label{eq:X}
\end{align}
where $H$ is the Heaviside function. The internal variable $\kappa$
is the \emph{branch point} where the cap begins, and $X(\kappa)$ is
the \emph{hydrostat intercept}: the current hydrostatic crush
strength. On the shear branch $I_1 > \kappa$, $F_c \equiv 1$; on the
compactive cap $X < I_1 \le \kappa$, $F_c$ falls quadratically from
one to zero, closing the surface at $I_1 = X$. At $I_1 = \kappa$
both $F_c = 1$ and $\partial F_c/\partial I_1 = 0$, so the surface is
$C^1$ across the branch point: there is no corner where cap meets
envelope. The strength at zero pressure is $A - C$; as compression
grows the exponential dies and strength saturates toward
$A - \theta I_1$. $R$ is the cap aspect ratio (its extent in $I_1$
per unit of $F_f$), and $N$ offsets the initial yield surface below
the failure envelope, leaving room for kinematic hardening.

\paragraph{Yield function and plastic potential.}
\begin{align}
  f(I_1, J_2^\xi, J_3^\xi, \backstress, \kappa)
    &= \Gamma^2 J_2^\xi - F_c(I_1,\kappa)\,\bigl(F_f(I_1) - N\bigr)^2,
    \label{eq:f}\\
  g(I_1, J_2^\xi, J_3^\xi, \backstress, \kappa)
    &= \Gamma^2 J_2^\xi - F_c^g(I_1,\kappa)\,\bigl(F_f^g(I_1) - N\bigr)^2,
    \label{eq:g}
\end{align}
with the plastic-potential analogues
\begin{align}
  F_f^g(I_1)  &= A - C \exp(L\,I_1) - \phi\,I_1, \label{eq:Ffg}\\
  F_c^g(I_1,\kappa) &=
    1 - H(\kappa - I_1)\,\frac{(I_1 - \kappa)^2}{[X^g(\kappa) - \kappa]^2},\\
  X^g(\kappa) &= \kappa - Q\,F_f(\kappa). \label{eq:Xg}
\end{align}
Non-associativity ($L \ne D$, $\phi \ne \theta$, $Q \ne R$) exists
because associated flow on a frictional envelope grossly
over-predicts dilatancy; flattening the potential's pressure
dependence ($\phi < \theta$, $L < D$) reduces the predicted plastic
volume expansion to measured levels. Associative flow corresponds to
$L = D$, $\phi = \theta$, $Q = R$.

\emph{Convention note.} In Eq.~\eqref{eq:Xg} the potential cap
position $X^g$ is built from the \emph{yield} failure function $F_f$
(parameters $D, \theta$), not from $F_f^g$: only the aspect ratio
$Q$ distinguishes the potential cap from the yield cap. This follows
\citet[Eq.~14]{Regueiro.Foster:2011} and the LAME reference
implementation. The distinction is invisible for associative
parameters and matters otherwise; the pre-2026 LCM code had it wrong
(Appendix~\ref{sec:history}).

\section{Flow Rule and Hardening Evolution}
\label{sec:flow}

The plastic strain rate follows the gradient of the plastic
potential \citep[Eq.~48]{Sun.Chen.Ostien:2014},
\begin{equation}
  \dot{\pstrain} \;=\; \dot\gamma\,\frac{\partial g}{\partial \stress},
  \label{eq:flow}
\end{equation}
where $\dot\gamma \ge 0$ is the consistency parameter.

\paragraph{Kinematic hardening.}
The deviatoric back stress evolves along the deviator of the flow
direction \citep[Eqs.~41--43]{Sun.Chen.Ostien:2014},
\begin{equation}
  \dot{\backstress} \;=\; \dot\gamma\,\halpha,
  \qquad
  \halpha \;=\; c^{\alpha}\,G^{\alpha}(\backstress)\,
               \dev\!\left(\frac{\partial g}{\partial \stress}\right),
  \qquad
  G^{\alpha} \;=\;
    \max\!\left(0,\; 1 - \frac{\sqrt{J_2^{\alpha}}}{N}\right),
  \label{eq:halpha}
\end{equation}
with $J_2^{\alpha} = \tfrac{1}{2}\backstress:\backstress$. The
limiter $G^\alpha$ drives the hardening rate to zero as the
translating yield surface approaches the fixed failure envelope,
which it reaches when $\sqrt{J_2^\alpha} = N$.

\emph{The clamp at zero in Eq.~\eqref{eq:halpha} is essential and
absent from the published papers.} The explicit update can overshoot
$\sqrt{J_2^\alpha}$ slightly past $N$; without the clamp,
$G^\alpha < 0$ \emph{reverses} the direction of kinematic hardening.
At a material point this is a small error, but at the structural
level it destabilizes the homogeneous solution: in LCM's verification
study a uniform-property confined-compression problem on a
$2 \times 2 \times 2$ mesh demonstrably bifurcated to a
non-homogeneous Newton solution at exactly the step where the back
stress reached its limit surface. The LAME reference implementation
has always floored its equivalent quantity (\texttt{GFUN}) at zero.
See Appendix~\ref{sec:history}.

\paragraph{Isotropic (cap) hardening.}
The cap position is driven by volumetric plastic strain through the
experimentally measured crush curve
\citep[Eqs.~44--46]{Sun.Chen.Ostien:2014},
\begin{equation}
  \dot\kappa \;=\; \dot\gamma\,h^{\kappa},
  \qquad
  h^{\kappa} \;=\;
    \frac{\trace(\partial g/\partial \stress)}
         {\partial \varepsilon_v^p/\partial \kappa},
  \qquad
  \varepsilon_v^p(\kappa)
    = W \Bigl[
        e^{\,D_1 \bar X - D_2 \bar X^2} - 1
      \Bigr],
  \quad \bar X = X(\kappa) - X(\kappa_0),
  \label{eq:kappa}
\end{equation}
where $\kappa_0$ is the initial cap branch point and
$W, D_1, D_2$ are material parameters. $W$ is the maximum
compactive volumetric plastic strain, essentially the initial
porosity: as $X \to -\infty$, $\varepsilon_v^p \to -W$ and all
porosity has been crushed out. The denominator in $h^\kappa$ is
evaluated by the chain rule
$(\partial\varepsilon_v^p/\partial X)(\mathrm{d}X/\mathrm{d}\kappa)$
in \texttt{compute\_dedkappa}.

\emph{Convention note.} The crush curve in Eq.~\eqref{eq:kappa} is
parametrized by the \emph{yield} cap position $X(\kappa)$ of
Eq.~\eqref{eq:X} --- parameters $R, D, \theta$ --- because $X$ is the
measurable hydrostatic elastic limit against which the curve is
calibrated. The LAME reference computes
$\mathrm{d}X/\mathrm{d}\kappa = 1 - R\,\mathrm{d}F_f/\mathrm{d}\kappa$
accordingly. The pre-2026 LCM code used the potential-surface
quantities ($Q, L, \phi$) here, which is wrong for non-associative
parameters (Appendix~\ref{sec:history}).

Two safeguards from the LAME reference, absent from the papers,
apply wherever $\kappa$ is updated:
\begin{itemize}
\item \emph{No cap contraction:} increments of $\kappa$ are clamped
      to $\Delta\kappa \le 0$ (LAME:
      \texttt{HK = MIN(HK,0)}, comment ``prevents cap
      contraction''). Dilation never re-opens porosity.
\item \emph{Cap lock at full compaction:} when
      $|\varepsilon_v^p(\kappa)| \ge W$ the crush curve is exhausted,
      its slope vanishes, and $h^\kappa$ as written diverges.
      Following the reference, $h^\kappa$ is then set to
      $-0.01\,K^2$ ($K$ the bulk modulus), which races the cap toward
      $-\infty$ so that pore-collapse plasticity shuts off rather
      than stalling against an immobile cap.
\end{itemize}

\section{Integration: Substepped Explicit Scheme}
\label{sec:integration}

The integrator is the refined explicit scheme of
\citet[Algorithms~1--2]{Sun.Chen.Ostien:2014}, upgraded with the
adaptive substepping of \citet{Sloan.Abbo.Sheng:2001}: within the
plastic part of a strain increment, modified-Euler (RK1/RK2) pairs
advance the state, the difference between the two estimates supplies
a local error measure that controls substep size, and every accepted
substep is followed by a drift correction back to the yield surface.

\paragraph{Rate evaluation.}
At state $\{\stress, \backstress, \kappa\}$, for a substrain
$\strainincr_{\mathrm{sub}}$, the consistency condition
$\dot f = 0$ gives
\begin{equation}
  \Delta\gamma \;=\;
  \frac{(\partial f/\partial \stress) : \Celast :
        \strainincr_{\mathrm{sub}}}{\chi},
  \qquad
  \chi \;=\; \tfrac{\partial f}{\partial \stress}
            :\Celast:\tfrac{\partial g}{\partial \stress}
          - \tfrac{\partial f}{\partial \backstress}:\halpha
          - \tfrac{\partial f}{\partial \kappa}\,h^{\kappa},
  \label{eq:dgamma}
\end{equation}
with $\partial f/\partial\backstress = -\partial f/\partial\stress$.
$\Delta\gamma$ is clamped at zero (elastic unloading within a
substep), and the increments are
\begin{equation}
  \Delta\stress = \Celast : \strainincr_{\mathrm{sub}}
                - \Delta\gamma\,\Celast : \tfrac{\partial g}{\partial\stress},
  \qquad
  \Delta\backstress = \Delta\gamma\,\halpha,
  \qquad
  \Delta\kappa = \min(0,\,\Delta\gamma\,h^{\kappa}).
  \label{eq:rates}
\end{equation}

\paragraph{Algorithm.}
With input state $\{\stress_n, \backstress_n, \kappa_n\}$ and strain
increment $\strainincr$:
\begin{enumerate}
\item \emph{Elastic trial} via~\eqref{eq:trial_small}. If
      $f(\stress^{\mathrm{tr}}, \backstress_n, \kappa_n) \le 0$,
      accept the trial and stop.
\item \emph{Substepping}, starting from the converged state (not the
      trial), with pseudo-time $T \in [0,1]$, $\Delta T = 1$
      initially, and $\Delta T_{\min} = 1/\texttt{Maximum Substeps}$.
      While $T < 1$:
      \begin{enumerate}
      \item Evaluate the increments \eqref{eq:rates} at the current
            state for $\strainincr_{\mathrm{sub}}
            = \Delta T\,\strainincr$ (RK1), and again at the
            RK1-advanced state (RK2).
      \item Form the modified-Euler state from the average of the two
            increment sets, and the relative error estimate
            \begin{equation}
              R \;=\; \frac{\|\Delta\stress_2 - \Delta\stress_1\|}
                           {2\,\max(\|\stress_{\mathrm{new}}\|,\,
                            \epsilon)} .
            \end{equation}
      \item If $R > \texttt{Substep Tolerance}$ and
            $\Delta T > \Delta T_{\min}$: reject, set
            $\Delta T \leftarrow
            \max\bigl(0.9\sqrt{\mathrm{STOL}/R},\,0.1\bigr)\,\Delta T$
            (floored at $\Delta T_{\min}$), retry.
      \item Otherwise accept: apply the drift correction below,
            advance $T \leftarrow T + \Delta T$, and grow
            $\Delta T \leftarrow
            \min\bigl(0.9\sqrt{\mathrm{STOL}/R},\,1.1\bigr)\,\Delta T$,
            capped at $1 - T$.
      \end{enumerate}
\item \emph{Plastic strain increment} recovered through elastic
      compliance,
      \begin{equation}
        \Delta\pstrain \;=\; (\Celast)^{-1} : (\stress^{\mathrm{tr}}
                                              - \stress_{n+1}),
      \end{equation}
      from which the outputs
      $\varepsilon_v^p \mathrel{+}= \trace\Delta\pstrain$ and
      $\varepsilon_{\mathrm{eqps}} \mathrel{+}=
      \sqrt{\tfrac{2}{3}}\,\|\dev\Delta\pstrain\|$ are accumulated.
\end{enumerate}

\paragraph{Drift correction.}
Explicit updates do not satisfy $f = 0$ exactly. After each accepted
substep, while $|f| > f_{\mathrm{tol}}$ and fewer than 20 iterations:
\begin{enumerate}
\item \emph{Consistent correction:} $\delta\gamma = f/\chi$ with all
      quantities at the current state; update
      $\stress \leftarrow \stress
       - \delta\gamma\,\Celast:\partial g/\partial\stress$,
      $\backstress \leftarrow \backstress + \delta\gamma\,\halpha$,
      $\kappa \leftarrow \kappa + \min(0, \delta\gamma\,h^\kappa)$.
\item \emph{Normal-correction fallback:} if the consistent correction
      increases $|f|$, discard it and instead project geometrically,
      \begin{equation}
        \delta\gamma \;=\;
          \frac{f}{\partial_\sigma f : \partial_\sigma f},
        \qquad
        \stress \leftarrow \stress
          - \delta\gamma\,\frac{\partial f}{\partial\stress},
      \end{equation}
      holding $\backstress, \kappa$ fixed. This is the
      \citet{Sloan.Abbo.Sheng:2001} form; the version printed in
      \citet{Sun.Chen.Ostien:2014} is dimensionally inconsistent
      (Section~\ref{sec:scope}).
\end{enumerate}
The yield function has units of stress squared, so the drift
tolerance is scaled to make the convergence test independent of the
unit system:
\begin{equation}
  f_{\mathrm{tol}} \;=\; 10^{-12}\,E^2 .
\end{equation}
(The historical absolute tolerance of $10^{-10}$ was unreachable in
Pa-based unit systems, so the corrector always ran to its iteration
cap.)

\paragraph{Branch consistency under AD.}
The substep control flow (acceptance, rejection, step growth)
branches on Sacado \emph{values}, so the Residual and Jacobian
evaluations of the same state take identical branch sequences and the
AD tangent differentiates exactly the code path that produced the
stress.

\paragraph{Evolving material parameters.}
The integrator accepts \emph{two} parameter sets, for the beginning and
end of the strain increment, and ramps linearly between them across the
substep pseudo-time: RK1 rates use the substep-start parameters, RK2
rates and the drift correction the substep-end parameters. With
constant parameters (the \texttt{Cap} model) the two sets are equal and
the scheme reduces to the fixed-parameter integrator. This mechanism
exists for models whose parameters follow an external field --- the
Permafrost model's ice saturation --- and it is not merely an accuracy
refinement. If instead the parameters jump once per step, a softening
step leaves the stored stress far outside the shrunken yield surface,
and the entire correction is delivered by the drift projection, whose
consistent/normal fallback decision then sits on a knife edge: in a
thaw-under-load verification study the resulting trajectories were
hypersensitive to roundoff (relative perturbations of $10^{-16}$ in the
moduli grew to order-one trajectory differences, with two distinct
attractors). Ramping the parameters keeps each substep's surface motion
small, the drift corrections small, and the trajectory robust --- the
same study then matched the reference implementation to
$\sim 10^{-15}$.

Two consistency requirements accompany evolving parameters. First, the
elastic strain, not the stress, is the state: when the moduli change
between steps, the stored stress must be re-expressed through the
current stiffness, $\stress_n \leftarrow \Celast(f) :
[\Celast(f_{\mathrm{old}})]^{-1} : \stress_n$, before integrating (the
finite-deformation path does this by construction, recovering the
elastic logarithmic strain from the stored plastic geometry). Second,
the previous step's parameter state must be available; the Permafrost
model stores its ice saturation as a state variable for this purpose.

\paragraph{Material parameters of the integrator.}
\texttt{Substep Tolerance} (STOL, default $10^{-4}$) and
\texttt{Maximum Substeps} (default 200). A rejected attempt counts
toward an overall attempt cap of $4\times$\texttt{Maximum Substeps}.

\paragraph{Hand-coded derivatives.}
The closed-form constitutive derivatives
$\partial f/\partial \stress$, $\partial g/\partial \stress$,
$\partial f/\partial \kappa$, $\halpha$, and
$\partial\varepsilon_v^p/\partial\kappa$ are provided in
\texttt{CapModel\_Def.hpp} (\texttt{compute\_dfdsigma},
\texttt{compute\_dgdsigma}, \texttt{compute\_dfdkappa},
\texttt{compute\_halpha}, \texttt{compute\_dedkappa}). The only
nontrivial tensor derivative among them is
\begin{equation}
  \frac{\partial J_3}{\partial \stress}
  \;=\; \sdev\cdot\sdev - \tfrac{2}{3} J_2\,\ident ,
  \label{eq:dJ3}
\end{equation}
a matrix product, \emph{not} the elementwise square of $\sdev$; the
two coincide only in principal axes (Appendix~\ref{sec:history}).
Every one of these closed forms is checked against complex-step
differentiation by the verification suite
(Section~\ref{sec:verification}).

\section{Finite-Deformation Kinematics}
\label{sec:finite}

With \texttt{Finite Deformation: true} the model uses the
multiplicative split and the isotropic logarithmic-strain /
Kirchhoff-stress framework
\citep{Simo:1992}: because every constitutive function in
Sections~\ref{sec:yield}--\ref{sec:flow} is an isotropic function of
its tensor arguments, the small-strain integrator of
Section~\ref{sec:integration} runs \emph{unchanged} with the
Kirchhoff stress $\kirch$ and the logarithmic elastic strain in the
slots that otherwise hold $\stress$ and $~\varepsilon$. Only the
kinematic pre- and post-processing differ. At each quadrature point,
with incoming $\Fdef$, stored $\Fplas_n$, and the deformation
gradient $\Fdef_n$ of the previous converged step:
\begin{enumerate}
\item \emph{Trial elastic left Cauchy--Green tensor} from the stored
      plastic configuration
      ($~C_p^{-1} = (\Fplas_n)^{-1}(\Fplas_n)^{-\mathrm{T}}$),
      \begin{equation}
        \bee^{\mathrm{tr}} \;=\; \Fdef\,~C_p^{-1}\,\transpose{\Fdef},
        \qquad
        \estrain^{\mathrm{tr}} \;=\;
        \tfrac{1}{2}\,\log_{\mathrm{sym}}\bee^{\mathrm{tr}},
      \end{equation}
      with $\log_{\mathrm{sym}}$ the principal (spectral) logarithm.
\item \emph{Starting stress and strain increment.} The elastic
      logarithmic strain at $t_n$ is recovered from the same plastic
      configuration and the old deformation gradient,
      \begin{equation}
        \estrain_n \;=\;
        \tfrac{1}{2}\,\log_{\mathrm{sym}}
        \bigl(\Fdef_n\,~C_p^{-1}\,\transpose{(\Fdef_n)}\bigr),
        \qquad
        \kirch_n = \Celast : \estrain_n,
        \qquad
        \strainincr = \estrain^{\mathrm{tr}} - \estrain_n .
      \end{equation}
      The integrator then starts from $\kirch_n$ with increment
      $\strainincr$, so its trial stress
      $\kirch_n + \Celast:\strainincr = \Celast:\estrain^{\mathrm{tr}}$
      is \emph{exact} for the hyperelastic logarithmic model: the
      incremental form introduces no approximation in the elastic
      relation.
\item \emph{Return mapping} of Section~\ref{sec:integration} on
      $\kirch$, yielding $\kirch_{n+1}, \backstress_{n+1},
      \kappa_{n+1}$.
\item \emph{Plastic configuration update.} The new elastic
      logarithmic strain follows from the returned stress,
      $\estrain_{n+1} = (\Celast)^{-1}:\kirch_{n+1}$, whence
      \begin{equation}
        \bee_{n+1} = \exp_{\mathrm{sym}}(2\,\estrain_{n+1}),
        \qquad
        ~C_{p,n+1}^{-1} = \Fdef^{-1}\,\bee_{n+1}\,\Fdef^{-\mathrm{T}},
        \qquad
        \Fplas_{n+1} = \bigl(~C_{p,n+1}\bigr)^{1/2},
      \end{equation}
      where the last expression is the symmetric positive-definite
      square root. The rotational part of $\Fplas$ is unobservable in
      an isotropic model, so the symmetric root is a legitimate
      representative; no exponential map of the flow direction is
      formed explicitly --- it is absorbed in the spectral
      $\log$/$\exp$ of $\bee$.
\item \emph{Cauchy stress} for assembly,
      $\stress_{n+1} = \kirch_{n+1}/J$ with $J = \det\Fdef$.
\end{enumerate}
The stored state in this mode is $\Fplas$ (identity-initialized)
plus the deformation gradient history that
\texttt{MechanicsProblem} already maintains; the small-strain mode
stores the strain tensor instead. All other state (stress, back
stress, $\kappa$, $\varepsilon_{\mathrm{eqps}}$, $\varepsilon_v^p$)
is common.

This wrapper is independent of how the inner return mapping is
integrated, which is why the desire for finite-deformation support
had no bearing on the decision to remove the implicit integrator.

\section{Tangents Are Not Hand-Coded: Automatic Differentiation}
\label{sec:ad}

\citet{Sun.Chen.Ostien:2014} do not document a consistent algorithmic
tangent $\partial \stress_{n+1}/\partial ~\varepsilon_{n+1}$, nor a
finite-deformation counterpart. LCM does not need either. LCM's
evaluators are templated on an evaluation trait; for
\texttt{PHAL\bcc AlbanyTraits\bcc Jacobian} every field entering the
kernel --- strain or deformation gradient, stored state, every
intermediate of the return mapping --- carries a Sacado forward-mode
AD type whose active directions are the element-level global degrees
of freedom. The entire integrator (trial, substepping, drift
correction, finite-deformation pre/post-processing) executes in AD
arithmetic, and the derivative of the output stress with respect to
every element DOF emerges as a by-product. Because the substep
control flow branches on Sacado values
(Section~\ref{sec:integration}), the differentiated path is exactly
the primal path. Sensitivities with respect to material parameters
(for calibration and inverse problems) are generated by the same
mechanism. The only hand-coded derivatives in the model are the
\emph{constitutive} ones listed in Section~\ref{sec:integration} ---
part of the model's definition, not of its integrator.

\section{Verification}
\label{sec:verification}

The implementation is verified, not merely regression-tested, by the
infrastructure in \texttt{tests/LCM/CapModelPlasticity3D}:

\paragraph{Independent reference implementation.}
\texttt{cap\_reference.py} is a deliberately separate transcription
of the model from the publications --- not a port of the C++. It
evaluates all stress gradients by complex-step differentiation
(machine-exact for the analytic branches), so the C++ closed-form
derivatives and the reference share no code and no derivation. Its
self-check verifies, among other identities, the $J_3$ derivative
\eqref{eq:dJ3} against complex step at a generic (non-principal-axes)
state, where the elementwise-square error would be $O(10^2)$ in
stress-squared units.

\paragraph{Material-point exactness of the test problems.}
Each verification problem is a single hexahedral element whose 24
degrees of freedom are all prescribed by face boundary conditions.
With no free degrees of freedom, the finite-element problem \emph{is}
the material point: no equilibrium iteration intervenes between the
prescribed strain history and the constitutive response, and the
comparison with the reference is exact rather than approximate. (The
importance of this is not academic: on a $2\times2\times2$ mesh the
same problem bifurcated away from the homogeneous solution when the
unclamped $G^\alpha$ defect of Appendix~\ref{sec:history} was
present.)

\paragraph{Load paths and results.}
Five paths are compared state-by-state (stress components, $\kappa$,
$\varepsilon_v^p$) over their full histories against the reference,
which runs the identical substepping algorithm:
\begin{center}
\small
\begin{tabular}{lp{3.4cm}p{4.4cm}l}
\toprule
Test & Path & Exercises & Max.\ rel.\ diff. \\
\midrule
\texttt{hydrostatic} & $~\varepsilon = -0.02\,t\,\ident$ &
  cap only; analytic identities & $\sim 10^{-14}$ \\
\texttt{confined} & $\varepsilon_{xx} = -0.04\,t$, 200 steps &
  shear--cap interaction, $\backstress$ to limit & $\sim 10^{-15}$ \\
\texttt{triaxial} & unequal normal strains &
  $\psi = 0.8$ (Lode), non-associative & $\sim 10^{-15}$ \\
\texttt{confined\_coarse} & confined, 20 steps &
  substep rejection/regrowth & $\sim 10^{-15}$ \\
\texttt{confined\_fd}, \texttt{triaxial\_fd} & as above, finite deformation &
  $\log$/$\exp$ kinematics & $\sim 10^{-12}$ \\
\bottomrule
\end{tabular}
\end{center}
The hydrostatic path additionally checks two analytic identities: on
the hydrostat the converged stress must satisfy $I_1 = X(\kappa)$
(observed to $\sim 10^{-13}$), and the accumulated
$\varepsilon_v^p$ must follow the crush curve of
Eq.~\eqref{eq:kappa} (observed to the per-step consistency error of
the discretization). The finite-deformation comparisons carry the
slightly larger $\sim 10^{-12}$ floor of the eigendecomposition
round-trips. The comparisons require \texttt{python3} with
\texttt{numpy} and \texttt{netCDF4} and are skipped with a status
message when these are unavailable.

\section{Material Parameters}
\label{sec:parameters}

All parameters are read from the material database with the names
below. The Salem limestone values are the associative set of
\citet[Table~1]{Sun.Chen.Ostien:2014}; stress-like quantities in MPa.

\begin{center}
\small
\begin{tabular}{llrp{6.4cm}}
\toprule
Input name & Symbol & Salem & Meaning \\
\midrule
\texttt{A}       & $A$        & 689.2    & saturated shear strength \\
\texttt{C}       & $C$        & 675.2    & strength deficit at zero pressure ($A{-}C$ = cohesion-like intercept) \\
\texttt{D}       & $D$        & 3.94e-4  & pressure scale of the exponential, yield surface \\
\texttt{theta}   & $\theta$   & 0.0      & linear pressure slope, yield surface \\
\texttt{N}       & $N$        & 6.0      & offset of initial yield below failure envelope \\
\texttt{R}       & $R$        & 28.0     & cap aspect ratio, yield surface \\
\texttt{kappa0}  & $\kappa_0$ & $-8.05$  & initial cap branch point \\
\texttt{W}       & $W$        & 0.08     & maximum compactive volumetric plastic strain ($\approx$ porosity) \\
\texttt{D1}      & $D_1$      & 1.47e-3  & crush-curve shape (linear) \\
\texttt{D2}      & $D_2$      & 0.0      & crush-curve shape (quadratic) \\
\texttt{calpha}  & $c^\alpha$ & 1.0e5    & kinematic hardening rate \\
\texttt{psi}     & $\psi$     & 1.0      & triaxial extension/compression strength ratio \\
\texttt{L}       & $L$        & 3.94e-4  & pressure scale, plastic potential \\
\texttt{phi}     & $\phi$     & 0.0      & linear pressure slope, plastic potential \\
\texttt{Q}       & $Q$        & 28.0     & cap aspect ratio, plastic potential \\
\midrule
\texttt{Substep Tolerance} & STOL & $10^{-4}$ &
  relative stress-error target per substep \\
\texttt{Maximum Substeps}  & ---  & 200 &
  smallest substep is $1/$this fraction of the increment \\
\texttt{Finite Deformation} & --- & false &
  $\log$/Kirchhoff kinematics of Section~\ref{sec:finite} \\
\bottomrule
\end{tabular}
\end{center}

For orientation with the Salem values: the initial hydrostatic crush
strength is $X_0 = \kappa_0 - R\,F_f(\kappa_0) \approx -460$~MPa,
i.e.\ pore collapse under pure pressure begins near
$p \approx 153$~MPa, while the cap branch point sits at only
$8$~MPa --- a long, flat cap, typical for porous rock.

\section{Calibration}
\label{sec:calibration}

The parameters of Section~\ref{sec:parameters} are obtained from a
small suite of standard rock-mechanics experiments. The groups below
are listed in the order in which they should be calibrated, because
later groups depend on earlier ones; the definitive fitting
procedures are documented in \citet[Section~4]{Fossum.Brannon:2004}.
All stress-like parameters must be supplied in one consistent unit
system ($D$, $L$, $D_1$ carry units of 1/stress, $D_2$ of
1/stress$^2$, $c^\alpha$ of stress; $W$, $\psi$, $R$, $Q$ are
dimensionless).

\paragraph{1.\ Elasticity ($E$, $\nu$).}
From \emph{unload--reload loops}, not initial loading slopes: first
loading of a geomaterial includes crack closure and seating, which
read as artificially soft moduli. The bulk modulus comes from
unloading slopes of the hydrostatic test (pressure vs.\ volumetric
strain), the shear modulus from unloading slopes of the triaxial
shear leg. Note that the original GeoModel uses pressure-dependent
elastic moduli while this implementation is linear elastic: calibrate
$(E,\nu)$ at the pressure range of the intended application, and do
not expect a single pair to span both the low- and high-confinement
regimes of the same rock.

\paragraph{2.\ Shear failure envelope ($A$, $C$, $D$, $\theta$).}
From \emph{triaxial compression (TXC) tests at several confining
pressures}: each test contributes one peak stress point
$(I_1, \sqrt{J_2})$, and the collection across pressures traces the
envelope \eqref{eq:Ff} for a least-squares fit. The parameters have
distinct anchors, which is what makes the fit well posed: $A - C$ is
the intercept at $I_1 = 0$, anchored by the unconfined compression
strength ($\sqrt{J_2} = \sigma_c/\sqrt{3}$ at $I_1 = -\sigma_c$);
$A$ is the high-pressure asymptote, anchored by the
highest-confinement tests; $1/D$ is the $I_1$ scale of the transition
between the two, requiring tests that span the curved part of the
envelope; and $\theta$ is a residual linear slope at high pressure
that should be held at zero unless the high-pressure data clearly
keep climbing --- it is strongly correlated with $A$ and $D$ when the
tested pressure range is narrow.

\paragraph{3.\ Lode strength ratio ($\psi$).}
The ratio of triaxial-extension to triaxial-compression strength at
the same mean stress, requiring \emph{triaxial extension} tests
(axial stress reduced below the confinement). These are harder to
run, so $\psi$ is often the first parameter sacrificed; $\psi = 1$
switches the Lode dependence off entirely. Convexity of the
Gudehus-type $\Gamma(\beta)$ in Eq.~\eqref{eq:Gamma} requires
$\psi \ge 7/9 \approx 0.778$; do not fit below that bound.

\paragraph{4.\ Cap and crush curve
($\kappa_0$, $W$, $D_1$, $D_2$, $R$).}
One \emph{hydrostatic crush test}, pressurized well past the elastic
limit with periodic unload--reload loops to separate elastic from
plastic volumetric strain, supplies most of this group. The
elastic-limit point gives $X_0$ ($I_1 = -3p$ at the first departure
from elastic response: the onset of pore collapse). Beyond it, the
stress state sits at $I_1 = X(\kappa)$ whenever it yields, so the
plot of plastic volumetric strain against $I_1$ \emph{is} the crush
curve of Eq.~\eqref{eq:kappa}: $W$ is its plateau --- the total
crushable volume fraction, which should agree with independently
measured porosity (for Salem limestone, $W = 0.08$ matches
$8\,\%$ porosity); $D_1$ sets the initial compaction rate; and $D_2$
is unidentifiable unless the measured curve is strongly
non-exponential --- keep it at zero otherwise.

The cap aspect ratio $R$ requires yield points on the cap \emph{away}
from the hydrostat: TXC tests at high confinement in which the
specimen compacts during shear (shear-enhanced compaction). The
locus of compaction-onset points in the $(I_1, \sqrt{J_2})$ plane
traces the cap ellipse, and $R$ is its width-to-height ratio.
Without such data, $R$ is commonly carried over from similar rock
(it varies less than the other parameters; roughly 15--30 for porous
rocks).

$\kappa_0$ is then \emph{derived, not measured}: invert
$X(\kappa_0) = \kappa_0 - R\,F_f(\kappa_0)$ for $\kappa_0$ given the
measured $X_0$ and the fitted $R$ and $F_f$. A common input error is
to place the measured crush pressure directly into \texttt{kappa0};
for Salem limestone the two differ by a factor of about 57
($\kappa_0 = -8.05$ vs.\ $X_0 \approx -460$~MPa).

\paragraph{5.\ Hardening ($N$, $c^\alpha$).}
Both come from the \emph{pre-peak portion} of the TXC stress--strain
curves --- the same tests as group~2, using the curve shape rather
than the peak. $N$ is the offset in $\sqrt{J_2}$ between initial
yield (departure from linearity, onset of acoustic emission, or first
unload--reload hysteresis) and peak; the model assumes this offset is
pressure independent, which should be checked across the test suite.
$c^\alpha$ is fit to the curvature of the hardening branch between
initial yield and peak: large values (Salem's $10^5$~MPa) drive the
yield surface to the failure envelope almost immediately, so the
response is nearly elastic--perfectly-plastic; small values give
gradual hardening. From monotonic data alone $N$ and $c^\alpha$ are
partially degenerate (a larger offset traversed faster resembles a
smaller offset traversed slower); they separate cleanly only with
load-reversal (cyclic) data, where the back stress reveals itself
through the reduced reverse-yield stress. If the application never
reverses the loading, the degeneracy is harmless.

\paragraph{6.\ Non-associative flow ($L$, $\phi$, $Q$).}
These parameters do not affect where yield occurs, only the direction
of plastic flow, so they are calibrated from \emph{plastic strain
ratios} rather than stress states. $L$ and $\phi$ control the
volumetric component of flow on the shear branch --- dilatancy:
measure the dilation rate
$\mathrm{d}\varepsilon_v^p/\mathrm{d}\varepsilon_q^p$ during
post-yield shear at several pressures and fit $L < D$,
$\phi < \theta$ so that $\trace(\partial g/\partial\stress)$ matches.
$Q$ does the same on the cap, from the ratio of compactive volumetric
to deviatoric plastic strain during shear-enhanced compaction. If
volumetric plastic-strain data are unavailable, set the flow
associative ($L = D$, $\phi = \theta$, $Q = R$) with the
understanding that dilatancy will be overpredicted. Recall from
Eq.~\eqref{eq:Xg} that $X^g = \kappa - Q\,F_f(\kappa)$ uses the
\emph{yield} $F_f$, so $Q$ is a pure aspect-ratio parameter --- it
does not pull $L$ or $\phi$ into the cap position.

\paragraph{Workflow and checks.}
The dependency order is: elasticity $\to$ failure envelope and $\psi$
$\to$ crush curve $\to$ $R$ $\to$ $\kappa_0$ (derived) $\to$
$N, c^\alpha$ $\to$ $L, \phi, Q$. Constraints worth verifying on any
fitted set: $A - C > N > 0$ (positive initial strength at zero
pressure); $D, D_1, W, R, Q > 0$; $\kappa_0 < 0$; and
$7/9 \le \psi \le 1$.

Two practical points for carrying this out in LCM. First, the
verification decks of Section~\ref{sec:verification} are
single-element, fully strain-controlled drivers:
\texttt{input\_hydrostatic.yaml} \emph{is} a hydrostatic crush
simulation and \texttt{input\_confined.yaml} \emph{is} an oedometer
test, so copying one and editing the boundary-condition ramp yields a
calibration driver that can be compared against each laboratory curve
directly, with no structural effects in the way. Second,
\texttt{Substep Tolerance} and \texttt{Maximum Substeps} are
numerical parameters that require no calibration and should be left
at their defaults; if results change when the tolerance is
tightened, the global time step was too coarse --- the remedy is not
in the material parameters.

\section{The Permafrost Model}
\label{sec:permafrost}

\texttt{Permafrost} (material-database name \texttt{Permafrost}) is
the cap-plasticity model specialized for frozen and thawing sediment
in the ACE Arctic coastal-erosion application
\citep{Frederick.etal:2021}, and replaces the \texttt{J2Erosion}
model there. \texttt{J2Erosion} extends $J_2$ plasticity with an
ice-saturation--dependent yield stress
($\sigma_0 = \sigma^{\mathrm{sed}} + f(T)\,
\sigma^{\mathrm{ice-sed}}$, with $f(T) \in [0,1]$ the ice saturation
from the sediment freezing curve) and, over time, accumulated five
failure criteria and several empirical adjustments (polynomial fits
of stiffness and strength in ice saturation and porosity,
ocean-exposure weakening factors, a peat-dependent strain limit).
\texttt{Permafrost} replaces those mechanisms with cap-model
quantities. The integration and constitutive functions are the
shared, verified integrator of Section~\ref{sec:integration}; the
kernel supplies parameters that vary per integration point with the
ice saturation.

\paragraph{Why the cap model fits permafrost.}
The eroding material spans a continuum from ice-bonded (frozen) to
thawed sediment, and the cap model contains both end members as
parameter limits. With $D \to 0$ the failure envelope flattens to
the pressure-insensitive strength $A - C$, and with $\psi = 1$ and
the cap pushed far out the model degenerates to von Mises --- the
\texttt{J2Erosion} ice model. At the other extreme, thawed sediment
is the material the cap model was designed for: frictional shear
strength plus compaction of a porous skeleton, neither of which
$J_2$ represents (in \texttt{J2Erosion} the thawed state is a
constant residual yield stress). Two further physical arguments:
frozen soils exhibit a closed strength envelope (strength rises with
confinement, peaks, and falls at high pressure through pressure
melting of pore ice), which the cap closure can represent
phenomenologically and a $J_2$ cylinder cannot; and thaw settlement
is compaction --- crush-curve physics, with $W$ supplied by the
effective-porosity profiles the ACE model already carries.

\paragraph{The saturation-to-parameter map.}
The material input carries two calibrated end-member parameter sets,
the \texttt{Frozen Parameters} and \texttt{Thawed Parameters}
sublists, and the ice saturation interpolates between them with one
rule per strength mechanism:
\begin{itemize}
\item \emph{Cohesion and bonding} ($A$, $C$, $N$, $c^\alpha$):
  linear in $f$ --- ice cementation, the direct analogue of the
  \texttt{J2Erosion} rule
  $c(f) = c^{\mathrm{sed}} + f\,c^{\mathrm{ice}}$.
\item \emph{Pore collapse} ($\kappa_0$, $W$, $D_1$): linear in $f$
  --- ice filling the pores resists crushing.
\item \emph{Friction and shape} ($D$, $\theta$, $L$, $\phi$, $R$,
  $Q$, $\psi$, $D_2$): properties of the sediment skeleton, taken
  from the thawed sublist and independent of $f$.
\item \emph{Elasticity}: the $(K, G)$ split described below.
\end{itemize}
The saturation itself comes from one of three sources: the
\texttt{ACE\_Ice\_Saturation} field computed by the coupled thermal
problem (\texttt{Have ACE Ice Saturation: true}), a time table
(\texttt{Ice Saturation Time Values} / \texttt{Ice Saturation
Values}), or the constant \texttt{Ice Saturation} (default 1.0,
frozen). The saturation seen at the previous converged step is kept
as the state \texttt{Ice\_Saturation\_State}, and the integrator
ramps the parameters from the previous to the current saturation
across its substeps --- the within-step ramping of
Section~\ref{sec:integration}, without which thaw-under-load
trajectories are hypersensitive to roundoff. Because the elastic
strain, not the stress, is the state, the stored stress is
re-expressed through the current stiffness when the moduli change
($\stress_n \leftarrow \Celast(f) : \Celast(f_{\mathrm{old}})^{-1}
: \stress_n$; the finite-deformation path does this by construction).
In the finite-deformation path the thermal stretch is removed from
$\Fdef$ before the mechanical response is evaluated.

\paragraph{Elasticity and ice saturation.}
\texttt{J2Erosion} scales the Young's modulus with ice saturation at
fixed Poisson's ratio, which scales $K$ and $G$ together. The
underlying mechanism is sound --- permafrost stiffness is governed by
what fills the pores and whether it bonds the grains, so the moduli
are properly functions of $f$ (temperature acts only through the
freezing curve) --- but the partition between the elastic constants
must differ. The thaw transition replaces pore ice
($G \approx 3.5$~GPa, $K \approx 9$~GPa) with pore water ($G = 0$,
$K \approx 2.2$~GPa) inside a mineral skeleton
($K_s \approx 36$~GPa for quartz), and the two moduli respond
entirely differently:
\begin{itemize}
\item \emph{Shear modulus:} the dominant, order-of-magnitude
      dependence. Frozen, ice welds the grain contacts and the
      composite $G$ is GPa-scale; thawed, water carries no shear and
      the unbonded skeleton's contact stiffness is tens of MPa --- a
      drop of roughly two orders of magnitude. This is why shear-wave
      velocity collapse is the standard geophysical signature of
      thaw.
\item \emph{Bulk modulus:} weak dependence, bounded below by the
      saturated mixture. While the pores remain water-filled, the
      undrained bulk stiffness follows the Reuss/Wood mixture of
      water and grains ($K \approx 5$~GPa at porosity $0.4$) --- a
      factor-of-two drop from frozen, not orders of magnitude. This
      is the classical Gassmann fluid-substitution result, with the
      additional loss of $G$ because ice, unlike water, had shear
      stiffness.
\item \emph{Poisson's ratio is then a prediction, not an input:}
      with $K \approx 5$~GPa and $G$ of tens of MPa, $\nu \to 0.499$
      --- thawed saturated soil approaches the undrained
      incompressible limit, up from $\nu \approx 0.3$ frozen.
\end{itemize}
Scaling $E$ at fixed $\nu$ instead predicts that the bulk stiffness
of thawed saturated sediment collapses along with the shear
stiffness, so thawed zones compact volumetrically under their own
weight far more than the pore water permits --- a plausible
contributor to the large displacements that motivated
\texttt{J2Erosion}'s displacement-norm failure criterion.

Accordingly, each end member carries its own $(K, G)$ pair in place
of modulus-multiplier fits. $G(f)$ interpolates geometrically
(log-linearly --- the range spans two orders of magnitude, so linear
interpolation in $f$ overweights the frozen value) and $K(f)$
linearly. The Wood-mixture bound is calibration guidance for the
thawed $K$, not a kernel computation: until pore pressure is
resolved (the $u$--$p$ extension below), the calibrated thawed
member uses drained-skeleton moduli, since an undrained $K$ would
make thawed mush bulk-stiffer than frozen soil. One numerical guard
is built in: the thawed saturated limit $\nu \to 0.499$
volumetrically locks low-order hexahedra, so the effective Poisson
ratio is capped (\texttt{Maximum Poissons Ratio}, default 0.45) by
reducing $K$ while preserving $G$ --- the trusted physics.

\emph{Design constraint:} environmental softening must not flow
through the hardening variable. When ice melts the physical cap
moves toward the origin, which would violate the
$\Delta\kappa \le 0$ clamp if it were imposed on $\kappa$. As with
temperature-dependent yield in metals, $\kappa$ stores only the
plastic compaction history while the surfaces are built from
$f$-dependent parameters evaluated fresh each step. A corollary for
calibration: because $\kappa$ is plastic memory while $\kappa_0(f)$
is a parameter, a state integrated under frozen parameters and then
read under thawed ones sits far inside the thawed crush curve ---
reinterpreted as heavily precompacted. End members should be
calibrated with a common $\kappa_0$ unless that reinterpretation is
intended.

\paragraph{Failure criteria.}
The cap model supplies constitutive failure indicators that replace
three of the five \texttt{J2Erosion} criteria with quantities that
have model meaning. Each indicator is normalized so that the value 1
means exhaustion of its mechanism, and each criterion is enabled by
a positive threshold (or limit) in the material input and disabled
at the default 0:
\begin{itemize}
\item \emph{Tension} (bit 0x01, \texttt{Tension Indicator
  Threshold}): the fraction of the zero-pressure shear capacity
  $F_f(0) - N$ lost to mean tension,
  $[\,(A - C) - F_f(I_1)\,]/(A - C - N)$, which reaches 1 where the
  shear surface pinches to the tensile apex $F_f(I_1) = N$ and the
  material has no remaining strength. This replaces the
  maximum-principal-stress check. The threshold \emph{must} be set
  below 1: because the yield function squares $F_f - N$, it has a
  spurious second branch beyond the apex on which the apparent
  strength grows again with tension. The integrator never reaches
  that branch in practice precisely because the element fails
  first --- the criterion is part of the model's domain of validity,
  not just an erosion trigger.
\item \emph{Backstress saturation} (bit 0x02, \texttt{Backstress
  Indicator Threshold}): $\sqrt{J_2^\alpha}/N$, equal to 1 when the
  kinematic hardening is exhausted ($G^\alpha = 0$: the yield
  surface has translated to the failure envelope and the shear
  response is perfectly plastic). This replaces the criterion that
  yield onset itself constitutes failure --- the cap model can
  distinguish yielded-but-hardening material from exhausted
  material, which $J_2$ cannot.
\item \emph{Crush exhaustion} (bit 0x04, \texttt{Crush Indicator
  Threshold}): $|\varepsilon_v^p(\kappa)|/W$ on the crush curve, the
  quantity whose approach to 1 engages the cap lock (full pore
  collapse).
\item \emph{Plastic distortion} (bit 0x20, \texttt{Maximum
  Equivalent Plastic Strain}): $\varepsilon_{\mathrm{eqps}}$ against
  a user limit, replacing the total-distortion limit.
\end{itemize}
The backstress and crush indicators approach 1 only asymptotically,
so their thresholds must also be below 1 (the implementation rejects
thresholds $\ge 1$). The two structural criteria --- tilt angle
(bit 0x08, \texttt{Critical Angle}) and displacement norm (bit 0x10,
\texttt{Maximum Displacement}), which detect toppled or departed
blocks --- are not constitutive and carry over from
\texttt{J2Erosion} unchanged, as does the element-death bookkeeping:
tripped modes accumulate in the per-point \texttt{failure\_modes}
bitmask (an STK-backed element state), the per-cell
\texttt{failure\_state} diagnostic encodes per-mode trip counts in
decimal digits, and a cell dies (\texttt{cell\_death} = 1, with live
propagation into the assembly's death-status vector so Newton stops
assembling condemned cells mid-step) when a user-settable number of
its integration points have failed
(\texttt{Failed Integration Points For Death}, default: all). All
criteria branch on Sacado values so the residual and Jacobian
evaluations take identical paths, and the death states are consumed
by the ACE solver and scatter machinery without modification.

\paragraph{The ACE environment.}
The kernel reads the same environmental data as \texttt{J2Erosion}.
Depth profiles against $z$ above mean sea level
(\texttt{ACE Z Depth File} with \texttt{ACE\_Porosity File} and
\texttt{ACE Peat File}, or the constant \texttt{ACE Bulk Porosity})
supply porosity and peat fraction at each integration point. The
porosity overrides the crushable volume $W$ --- the pore space is
the crushable volume, independent of what fills it --- and peat
raises the eqps limit by the factor $(1 + \mathrm{peat})$, a
provisional parity rule pending calibration. Sea level against time
(\texttt{ACE Sea Level File} / \texttt{ACE Time File}) drives
ocean-exposure weakening of submerged cells in the erodible sets:
the cohesion by its factor ($C$ rises toward $A$, shrinking the
zero-pressure strength $A - C$ and the offset $N$ while leaving the
friction slope and asymptote intact), the moduli uniformly (so
$\nu$, and the stored-stress re-expression, are unaffected), and the
eqps limit by its factor. Unlike the saturation, the weakening
switches on instantaneously when a cell submerges (the
\texttt{J2Erosion} convention); if waterline flips ever destabilize
the drift correction the way unramped parameter jumps do, the
submersion state should be ramped like the saturation.

\paragraph{Material input.}
In addition to the cap parameters of Section~\ref{sec:parameters}
(split per the map above between the \texttt{Frozen Parameters} and
\texttt{Thawed Parameters} sublists, with the friction/shape
parameters read from the thawed sublist), the \texttt{Permafrost}
material block reads:

\begin{center}
\small
\begin{tabular}{lrp{6.6cm}}
\toprule
Input name & Default & Meaning \\
\midrule
\texttt{K}, \texttt{G} (per sublist) & --- &
  end-member bulk and shear moduli \\
\texttt{Maximum Poissons Ratio} & 0.45 &
  cap on the effective $\nu$, preserving $G$ \\
\midrule
\texttt{Have ACE Ice Saturation} & false &
  read $f$ from the coupled thermal field \\
\texttt{Ice Saturation Time Values}/\texttt{Values} & --- &
  time-table source for $f$ \\
\texttt{Ice Saturation} & 1.0 & constant source for $f$ \\
\midrule
\texttt{Tension Indicator Threshold}    & 0 (off) & see failure criteria \\
\texttt{Backstress Indicator Threshold} & 0 (off) & '' \\
\texttt{Crush Indicator Threshold}      & 0 (off) & '' \\
\texttt{Maximum Equivalent Plastic Strain} & 0 (off) & '' \\
\texttt{Critical Angle}                 & 0 (off) & tilt, radians \\
\texttt{Maximum Displacement}           & 0 (off) & displacement norm \\
\texttt{Failed Integration Points For Death} & 0 (= all) &
  per-cell death threshold \\
\texttt{Disable Erosion} & false & criteria never trip \\
\midrule
\texttt{ACE Z Depth File} & --- & $z$ profile abscissae \\
\texttt{ACE\_Porosity File}, \texttt{ACE Peat File} & --- &
  depth profiles \\
\texttt{ACE Bulk Porosity} & 0 (off) & constant porosity ($W$ override) \\
\texttt{ACE Sea Level File}, \texttt{ACE Time File} & --- &
  sea level against time \\
\texttt{ACE Cohesion Weakening Factor}   & 1 & submerged-cell weakening \\
\texttt{ACE Stiffness Weakening Factor}  & 1 & '' \\
\texttt{ACE Eqps Limit Weakening Factor} & 1 & '' \\
\midrule
\texttt{Finite Deformation}, \texttt{Substep Tolerance},
\texttt{Maximum Substeps} & &
  as in Section~\ref{sec:parameters} \\
\bottomrule
\end{tabular}
\end{center}

Per-criterion indicator fields, the failure bookkeeping states, and
the standard cap states are available for Exodus output through the
usual \texttt{Output ...} flags.

\paragraph{Calibration.}
The thawed end member comes from standard soil-mechanics cap
parameters for silty/clay sediments plus the ACE porosity profiles
($W$); the frozen end member from refitting the experimental data
behind the \texttt{J2Erosion} stiffness/strength fits to cap
parameters, with friction ($\theta$, $D$) taken from the thawed
sediment so the frozen fit reduces to cohesion scaling --- the
original \texttt{J2Erosion} assumption, now in its natural place.
The script
\texttt{tests/LCM/ACE/MiniErosionPermafrost/calibrate\_permafrost.py}
performs this conversion: it evaluates the \texttt{J2Erosion} fit
polynomials (which encode the experimental program) at the end
members for a given porosity and deck scales and emits the
\texttt{Frozen}/\texttt{Thawed Parameters} sublists, including the
mapping of the distortion strain limit to an eqps limit. The
single-element verification decks of Section~\ref{sec:verification}
double as calibration drivers.

\paragraph{Verification.}
The \texttt{Permafrost} battery extends the methodology of
Section~\ref{sec:verification}, with the reference implementation
mirroring the saturation map, the parameter ramping, and the failure
indicators. Trajectory tests: equivalence with \texttt{Cap} at
$f = 1$, the full map at constant $f = 0.5$ (the $\nu$ cap active),
and thaw under load ($f: 1 \to 0.3$ during confined compression) all
match at machine precision. The porosity-profile test drives the two
Gauss-point depth groups of a single element as independent material
points against two reference runs. Five per-mode death tests verify
each failure criterion: the indicator series match the reference at
machine precision and the
\texttt{failure\_modes}/\texttt{failure\_state}/\texttt{cell\_death}
series match the reference-predicted trip step exactly (the
displacement criterion additionally checks the per-point indicator
multiset against the analytic Gauss-point displacements). Finally, a
MiniErosion-style coupled test runs \texttt{Permafrost} under the
ACE thermo-mechanical solver: a deliberately tight displacement
limit kills the settled upper cells early, thaw-driven failures
follow as the bluff warms, and the run continues on the eroded mesh
--- exercising live death, the active-part update, and the workset
rebuilds on the shared mesh.

\paragraph{Future extensions.}
In order of expected payoff: rate sensitivity by the Duvaut--Lions
wrap of \citet{Regueiro.Foster:2011}, which regularizes any inviscid
integrator and would represent storm-timescale ice creep; and
pore-pressure/effective-stress coupling --- the source paper
\citet{Sun.Chen.Ostien:2014} is itself a $u$--$p$ poromechanics
formulation, so the model is built for that extension if ACE ever
resolves pore fluid. The $u$--$p$ extension would also let the
thawed bulk modulus take its undrained (Wood-bound) value instead of
the drained-skeleton compromise above.

\appendix

\section{Revision History of the Formulation}
\label{sec:history}

In 2026 the implementation was audited line-by-line against the
sources listed in Section~\ref{sec:scope}, treating the code as
unreliable, and then verified against the independent reference of
Section~\ref{sec:verification}. Results produced with the pre-audit
code differ from the current model in the following ways; all four
defects are absent from every published validation case, which is
why they survived.

\begin{enumerate}
\item \emph{$\partial J_3/\partial\stress$ used the elementwise
      square} $s_{ij} s_{ij}$ (no sum) instead of the matrix product
      $(\sdev\cdot\sdev)_{ij}$ in Eq.~\eqref{eq:dJ3}. The two agree
      in principal axes and the term is multiplied by
      $\mathrm{d}\Gamma/\mathrm{d}J_3 \propto (1 - 1/\psi)$, so any
      run with $\psi = 1$ --- including all of
      \citet{Sun.Chen.Ostien:2014} --- never exercised it. Runs with
      $\psi \ne 1$ received wrong plastic flow directions off the
      principal axes.
\item \emph{$X^g$ was built from $F_f^g$} ($L, \phi$) instead of the
      yield $F_f$ ($D, \theta$) required by
      \citet[Eq.~14]{Regueiro.Foster:2011}; see Eq.~\eqref{eq:Xg}.
      Invisible for associative parameters.
\item \emph{The crush-curve derivative used the potential-surface
      cap position} ($Q, L, \phi$) instead of the yield $X(\kappa)$
      ($R, D, \theta$); see Eq.~\eqref{eq:kappa}. Invisible for
      associative parameters.
\item \emph{$G^\alpha$ was unclamped}, allowing hardening reversal
      after overshoot of the back-stress limit surface; see
      Eq.~\eqref{eq:halpha}. This is the likely root cause of the
      Newton divergence that led to the disabling of the original
      \texttt{CapModelPlasticity3D} regression tests years earlier.
\end{enumerate}

In the same effort: the drift tolerance was made unit-system
invariant ($10^{-12} E^2$); the $\Delta\kappa \le 0$ clamp, present
in the code but undocumented, was confirmed against the LAME
reference and documented; the cap lock at full compaction was added
from the reference; adaptive substepping was added; the implicit
variant (\texttt{CapImplicit}) was removed; the model was renamed
from \texttt{CapExplicit} to \texttt{CapModel} with material-database
name \texttt{Cap}; and the earlier finite-deformation path (which
skipped the predictor, disabled the normal-correction fallback, and
updated $\Fplas$ by an exponential map of the final flow direction)
was replaced by the formulation of Section~\ref{sec:finite}.

\begin{thebibliography}{9}

\bibitem[Frederick et~al.(2021)Frederick, Mota, Tezaur, and
  Bull]{Frederick.etal:2021}
J.~Frederick, A.~Mota, I.~Tezaur, and D.~Bull.
\newblock A thermo-mechanical terrestrial model of {Arctic} coastal erosion.
\newblock {\em Journal of Computational and Applied Mathematics}, 397:113533,
  2021.

\bibitem[Fossum and Brannon(2004)]{Fossum.Brannon:2004}
A.~F. Fossum and R.~M. Brannon.
\newblock {\em The Sandia GeoModel: Theory and User's Guide}.
\newblock Sandia Technical Report SAND2004-3226. Sandia National Laboratories,
  2004.

\bibitem[Fossum and Fredrich(2000)]{Fossum.Fredrich:2000}
A.~F. Fossum and J.~T. Fredrich.
\newblock {\em Cap plasticity models and compactive and dilatant pre-failure
  deformation}.
\newblock Sandia Technical Report SAND2000-1056. Sandia National Laboratories,
  2000.

\bibitem[Foster et~al.(2005)Foster, Regueiro, Fossum, and
  Borja]{Foster.etal:2005}
C.~D. Foster, R.~A. Regueiro, A.~F. Fossum, and R.~I. Borja.
\newblock Implicit numerical integration of a three-invariant,
  isotropic/kinematic hardening cap plasticity model for geomaterials.
\newblock {\em Computer Methods in Applied Mechanics and Engineering},
  194:5109--5138, 2005.

\bibitem[Regueiro and Foster(2011)]{Regueiro.Foster:2011}
R.~A. Regueiro and C.~D. Foster.
\newblock Bifurcation analysis for a rate-sensitive, non-associative,
  three-invariant, isotropic/kinematic hardening cap plasticity model for
  geomaterials: Part I. Small strain.
\newblock {\em International Journal for Numerical and Analytical Methods in
  Geomechanics}, 35:201--225, 2011.

\bibitem[Simo(1992)]{Simo:1992}
J.~C. Simo.
\newblock Algorithms for static and dynamic multiplicative plasticity that
  preserve the classical return mapping schemes of the infinitesimal theory.
\newblock {\em Computer Methods in Applied Mechanics and Engineering},
  99(1):61--112, 1992.

\bibitem[Sloan et~al.(2001)Sloan, Abbo, and Sheng]{Sloan.Abbo.Sheng:2001}
S.~W. Sloan, A.~J. Abbo, and D.~Sheng.
\newblock Refined explicit integration of elastoplastic models with automatic
  error control.
\newblock {\em Engineering Computations}, 18(1/2):121--154, 2001.

\bibitem[Sun et~al.(2014)Sun, Chen, and Ostien]{Sun.Chen.Ostien:2014}
W.~Sun, Q.~Chen, and J.~T. Ostien.
\newblock Modeling the hydro-mechanical responses of strip and circular punch
  loadings on water-saturated collapsible geomaterials.
\newblock {\em Acta Geotechnica}, 9(6):903--934, 2014.
\newblock doi:10.1007/s11440-013-0276-x.

\end{thebibliography}

\end{document}
